Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2022-01-27 analysis/paper/paper2.Rmd
Version with SNV performed before detrending for raman

@@

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_baseline_corrected_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

@@

Points.raman <-
  raman.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
           type == "Point" | type == "Point fragment" | type == "Preform")

#Fill the NA fields in the munsell_hue column and mark them as colourless samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy. The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89]. Previous research suggests a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/forskning/infrastruktur/mal/]. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice, on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.

@@

@@

```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra. The main explanatory power can be found in PC 1, which groups the data based on the quartz and D-G bands. As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA. PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the bottom left, that may result from Al, Si and OH vibrations."}

#perform PCA with SNV normalization and mean-center
raman.pca <-
  prcomp(standardNormalVariate(X = Points.raman[,c(30:321)]), center = TRUE, scale = FALSE)

@@

basic_plot2 <-
  fviz_pca_ind(raman.pca, axes = c(1,4), label="none")

#prepare labels for PCs
raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
raman.pc4var <- round(summary(raman.pca)$importance[2,4]*100, digits=1)
raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
raman.pc4lab <- paste0("PC4 (",as.character(raman.pc4var),"%)")

@@

comb +
  plot_layout(nrow = 2, guides = "collect")
```

>>>>

#Import descriptive metadata
metadata.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))

@@

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220127.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

@@

#Fill the NA fields in the munsell_hue column and mark them as colourless samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"

Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman scattering and XRF. The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89]. Previous research suggests a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/forskning/infrastruktur/mal/]. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.

@@

```{r raman-prep-data}
#aggregate observations by group(sample) and calculate average of wavelength measurements
raman.averaged <- 
  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)

#perform Standard Normal Variate on raman data
raman.averaged[,2:466] <- 
  standardNormalVariate(raman.averaged[,2:466])

#merge data with metadata
#raman_merged <- 
#  as.data.frame(merge(metadata, raman_averaged, by='sample_id'))

#select sample_id and data columns and transpose in prep for baseline de-trend
raman.transposed <- 
  raman.averaged %>% 
    select(sample_id, `92.88`:`2503.59`)

raman.transposed <- 
  t(raman.transposed)

#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
#ChemoSpec requires data to be read from file
write.table(raman.transposed, file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )

#prepare vector with sample_id for reading raman data into spectra object
sample_id <- 
  raman.averaged$sample_id

#read transposed raman data into spectra object
raman <-matrix2SpectraObject(
  gr.crit = sample_id,
  gr.cols = c("auto"),
  freq.unit ="Wavelength cm-1",
  int.unit ="Intensity",
  descrip ="Bifacial points measurements",
  in.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv",
  sep = ";",
  dec = ".",
  chk = TRUE,
  out.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_spec_object")

#call on relevant baseline method (modified polynomial fitting) from baseline package and 
#return corrected spectra to spectra object
raman.baseline <- 
  baselineSpectra(raman,
                  int = FALSE,
                  method = "modpolyfit",
                  retC = TRUE)

#spectra are stored in "Spectra" object, not a data frame
#need to extract samples and frequencies from that object 
raman.tmp <- 
  cbind(raman.baseline$freq, t(raman.baseline$data))
#transpose data to get frequencies as columns
raman.tmp <- 
  as.data.frame(t(raman.tmp))
#Add sample_id column and remove the initial "X" character introduced to sample names by R
raman.tmp$sample_id <- 
  c("sample_id", sub("^X*", "", raman.baseline$names))

#Make first row headers and remove the first row with the names
names(raman.tmp) <-
  raman.tmp[1,]
raman.tmp <- 
  raman.tmp[-1,]

#merge raman de-trended data with metadata
raman.baseline.merged <- 
  as.data.frame(merge(metadata.csv, raman.tmp, by='sample_id'))

#select and filter raman data to focus on relevant material and artefact types
Points.raman <-
  raman.baseline.merged %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
```

@@

```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra. The main explanatory power can be found in PC 1, which groups the data based on the quartz and D-G bands. As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA. PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the bottom left, that may result from Al, Si and OH vibrations."}
#perform PCA with SNV normalization and mean-center on the 
raman.pca <-
  prcomp(Points.raman[,c(30:321)], center = TRUE, scale = FALSE)

@@

basic_plot2 <-
  fviz_pca_ind(raman.pca, axes = c(2,3), label="none")

#prepare labels for PCs
raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
raman.pc4var <- round(summary(raman.pca)$importance[2,4]*100, digits=1)
raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
raman.pc4lab <- paste0("PC3 (",as.character(raman.pc4var),"%)")

@@

comb +
  plot_layout(nrow = 2, guides = "collect")
